Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-08-20 manuscript.Rnw
Combined both explained variance tables into main text

% Notes on Poorter et al. 2015.
% - we have to demonstrate that our work is still novel
% - their focus is firstly on the size-related shifts in biomass partitioning, we don't study that in detail, and can now justify that by referring back to Poorter.
% - they don't have SLA data for the same plants, and in fact predict what we found: 
% "It is also known that, on average, the specific
% leaf area (SLA; leaf area per unit leaf mass) of evergreen woody
% species is 2–3 times higher than that of deciduous species
% (Poorter et al., 2009). Hence, all else being equal, the 2–3 times
% lower SLA in evergreens would have to be compensated by a 2–3
% time higher LMF to arrive at the more or less similar LAI."
% - so basically we can say in the intro that we wish to test this hypothesis (as we already say), and also that Poorter speculated that this may be so (although here they contrast deciduous vs. evergreen, and not angiosperm vs. gymnosperm, so our results are still novel in that case as well)
% - we have to clearly state that our data has *no overlap* whatsoever with Poorter, this is a good thing.
% - their data includes potted plants in glass houses and field-grown, as well as stand biomass data divided by stocking. Our data are more homogenous, with only field-grown plants, individual estimates.
% - we have larger trees included; in fact the largest in the world, we need to state that in the intro to again show the novelty (our max mass is ca. 300t, theirs is 1t)
% - they don't actually show LMF vs. plant size split by PFT (their Fig 2a), so our result is also not scooped there.
% - they found no difference in the LMF residual between evergreen angiosperm and gymnosperm (Fig. 4), or at least I don't think so.

@@

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes. While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are only constrained, due to limited available data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.

% This para feels out of place. Perhaps better incorporated as part of material for last para in intro? Or towards the end in the first paragraph of the discussion?
Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning. Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability. Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs. This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.

@@

Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale. The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig. S1), across the three major biomes (boreal, temperate and tropical). In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height). One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}. We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals. In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. For each analysis, we used different subsets of the data because not all variables were measured in each study. Sample sizes by PFT are summarized in Tables S2-S5. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for methods details for each contributed study). For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}, with  different conversion factors applied to pines (\emph{Pinus} spp.) and non-pines. This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model, as follows.

@@

>>>>

% Notes on Poorter et al. 2015.
% - we have to demonstrate that our work is still novel
% - their focus is firstly on the size-related shifts in biomass partitioning, we don't study that in detail, and can now justify that by referring back to Poorter.
% - they don't have SLA data for the same plants, and in fact predict what we found: 
% "It is also known that, on average, the specific
% leaf area (SLA; leaf area per unit leaf mass) of evergreen woody
% species is 2–3 times higher than that of deciduous species
% (Poorter et al., 2009). Hence, all else being equal, the 2–3 times
% lower SLA in evergreens would have to be compensated by a 2–3
% time higher LMF to arrive at the more or less similar LAI."
% - so basically we can say in the intro that we wish to test this hypothesis (as we already say), and also that Poorter speculated that this may be so (although here they contrast deciduous vs. evergreen, and not angiosperm vs. gymnosperm, so our results are still novel in that case as well)

% - we have to clearly state that our data has *no overlap* whatsoever with Poorter, this is a good thing.
% - their data includes potted plants in glass houses and field-grown, as well as stand biomass data divided by stocking. Our data are more homogenous, with only field-grown plants, individual estimates.
% - we have larger trees included; in fact the largest in the world, we need to state that in the intro to again show the novelty (our max mass is ca. 300t, theirs is 1t)
% - they don't actually show LMF vs. plant size split by PFT (their Fig 2a), so our result is also not scooped there.
% - they found no difference in the LMF residual between evergreen angiosperm and gymnosperm (Fig. 4), or at least I don't think so.

@@

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes. While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.

@@

Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning. Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}. Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs. This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations. Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems could be related to differences in leaf mass per area. Here we can directly test this hypothesis as our database includes many observations of leaf area as well as leaf mass measured on the same plants.

A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale. The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig. S1), across the three major biomes (boreal, temperate and tropical). In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height). One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}. We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals. The database does not overlap with the recent compilation of \citet{poorter2015}. In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. For each analysis, we used different subsets of the data because not all variables were measured in each study. Sample sizes by PFT are summarized in Table 1. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for details on the methods for each contributed study). For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}, with  different conversion factors applied to pines (\emph{Pinus} spp.) and non-pines. This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model, as follows.

@@

<<results='asis'>>=
tab1 <- table_varpart2
tab2 <- table_gamr2MATMAP

tidytab <- function(x, labeltopleft=""){
  names(x)[1] <- "Variable"
  x[,1] <- paste0(x[,1],"(",x[,2],")")
  x[,2] <- NULL
  x[,2:ncol(x)] <- round(x[,2:ncol(x)], 2)  
  x <- unname(rbind(names(x),as.matrix(x)))  
  x <- rbind(c(labeltopleft,"","Predictors",""),x)
return(x)
}

tab1 <- tidytab(tab1, "Mixed-effects model")
tab2 <- tidytab(tab2, "GAM")

tab <- rbind(tab1, tab2)

print(xtable(tab, align="lrrrr",
             caption="Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables. Each variable was added to a linear mixed-effects model, using species within dataset as a random effect. The $R^2$ shown is that explained by the fixed effects only. All fixed effects were highly significant (P \\textless 0.001). In addition to the fixed effects shown, all interactions were added to each of the models."),
             hline.after=c(0,1,7,8,14),
             include.rownames=FALSE, 
             include.colnames=FALSE, 
             caption.placement="top", sanitize.text.function=function(x){x})
@
